Work Fragmentation in Developer Interaction Data

### 2015-08-07 paper.tex
section 10 and 11

\subsection{RQ3: Local relationships between interruptions and productivity}
The sessions with UDC went through the same process of transformation into 10 chunks of equal time. The results on the Figure \ref{fig:bins_int_udc} shows similar patterns to the ones we observed in Mylyn.
\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_without_int_med} \\
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_with_int1_med} \\
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_with_int2_med} 
\end{tabular}
\caption{Global representation of a session with: no interruptions (top); only short interruptions (middle); and at least one prolonged interruption (bottom). }
\label{fig:bins_int_udc}
\end{figure}

The median of editions and selections per minute are grater in sessions without interruptions, and the editions are more common than selections. When facing at least one prolonged interruption we can see the metrics drop in comparison with short interruptions and the selections are now more frequent than the editions events. This agrees with the results with Mylyn on this point and in the fact that during the beginning and the end of a session both metrics are lower than when in the middle of the session, but in this case the effect is less pronounced.

On a local analysis around the interruptions, we can see  interesting behaviors of the edition and selections metrics in a similar manner as with Mylyn. In the Figure \ref{fig:all_ints_udc} we show the frequency of editions and selections around the interruptions, specifically 30 minutes before and after. We can see that close to the interruption the frequency of editions and selections drastically drop, and in a radius of around 10 minutes both metrics are below the overall median. This phenomenon is also visible on the same version of this plot with Mylyn, agreeing that the activity of programming gradually decreases before the interruption and takes 5 minutes or more after it to restore the intensity of work above the median.

@@

After this, we separated the interruptions in three categories: those with positive effects afterwards, those a with negative effect and those with no clear effect. Both positive and negative interruptions have a proportion of $44\%$ each of all the interruptions, and the neutrals stand with the $12\%$. The results are on the Table \ref{tbl:local_effect_udc}.

@@

The table \ref{tbl:local_effect_udc} show a similar pattern on negative, positive and neutral interruptions in comparison with the results on Mylyn. After a positive interruption both metrics tend to increase at a high rhythm. In contrast, when facing the effects of a negative interruption the metrics take longer to reach the median.

\section{RQ4: Recovery Time after an Interruption}
\subsection{Edition and selections during recovery time}
The amount of information we have with Usage Data Collector allows to analyze more specific sections on a programming session. We further investigated the immediate activities after an interruption that we name the recovery time, which is a time period taken by the programmer to resume the interrupted work.

@@

From those valid interruptions, we split the time after the interruption event into two phases: the first 15 minutes labeled as \textit{Recovery Time} and the rest as \textit{After Recovery Time}. We calculated the metrics of every of both phases, after diving the Recovery Time into chinks of five minutes to better visualize the evolution of the metrics.

Considering the results obtained on the local analysis of an interruption, we can conclude that before an interruption it's common to have a reduction of both editions and selections per minute. Therefore, the \textit{After Recovery Time} time was split into two phases as well: one that takes the time where the metrics drop which happens 10 minutes before the next interruption (labeled as  \textit{Before Interruption}), and the rest is the peak of the session after the Recovery Time. The duration of the phase \textit{Before Interruption} was chosen with the conclusions after the local analysis around the interruptions.  We can see on the Figure \ref{fig:box_recovery_udc} the evolution of the metrics through the phases.

@@

The three metrics show a tendency to increase and reach a maximum between the last phase of the \textit{Recovery Time} and the peak of the session. Moreover, the fall of the productivity on the last minutes before the next interruptions agrees with the observations on the local analysis of the time around an interruption.

@@

We can see that after a positive interruption the edit ratio during recovery time is much higher than after a negative interruption, which agrees with the results on the local analysis of interruptions and the effects on the metrics. The reason of this change is probably due to the difference on the activities performed after the interruption according to the effect

@@

To better understand the reasons of the difference between positive and negative interruptions, we did an analysis of the common events during Recovery Time, depending on the nature of the interruption. For this we created a new classification of events on a lower level that takes into consideration what kind of activity is performed by the programmer. The new classification contains the categories described on the Table \ref{tbl:new_events_udc}.

\begin{table}[ht!]
\renewcommand{\arraystretch}{1.3}
\caption{New category of events. }
\label{tbl:new_events_udc}
\centering
\begin{tabular}{|p{1.7cm}|p{4 cm}|} 
  \hline 
category & description \\  
  \hline 
    \hline 
clean-build &  Events that clean, build or run a project \\
control & For control version events \\
debug & Events executed during debugging sessions  \\
edit-text & Text edition events  \\
file & Management of files (saving, closing, opening, etc.)  \\
find-replace & Search and replacement of text  \\
high-nav & Navigation through classes, windows and artifacts \\
perspective & Change of perspective on Eclipse  \\
refactoring & For refactoring events \\
search & Search of text  \\
text-nav & Navigation on the current text \\

@@

We calculated a weighted average of every event based on the average frequency of execution during the first phase of Recovery Time after a positive or negative interruption, multiplied by the percentage of users that make use of that event. The same calculation was made using statistics considering all the times with no distinction. Then we compared the weighted average changes to have evidence of what kind of events are executed after an interruption depending on its effect. The statistics can be found on the Table \ref{tbl:stats_events}.

@@

We can conclude that after a positive interruption there are more assertive actions with a clear intention like refactoring a class or saving all changes. And after a negative interruption there are more uncertainty and doubt judging by the activities performed like debugging, searching for text or navigating around classes.

It is noticeable the relation between type of events. For example is natural to think that if there is a lot of debugging events there is also a lot of clean and build events as they are necessary to begin a debugging session. There is also a link between high level navigation and debugging because navigating into a method may take you to another class.
\section{Discussion}
Although they represent the same kind of information both Mylyn and UDC datasets have some peculiarities that make them unique and require different treatments.

>>>>

\subsection{RQ3: Local relationships between interruptions and productivity}
The sessions with UDC went through the same process of transformation into 10 chunks of equal time. The results on the Figure \ref{fig:bins_int_udc} shows similar, but not identical patterns compared to the ones we observed in Mylyn.
\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_without_int_med} \\
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_with_int1_med} \\
\includegraphics[width=.80\linewidth,clip=, angle=0]{figures/udc_with_int2_med} 
\end{tabular}
\caption{Global representation of a session with: no interruptions (top); only short interruptions (middle); and at least one prolonged interruption (bottom). }
\label{fig:bins_int_udc}
\end{figure}

The median of editions and selections per minute are greater in sessions without interruptions, and the editions are more common than selections. Sessions with short interruptions only show level of edit activity similar to selection activity overall, but with a tendency to rise overtime (except a drop in the last segment). When facing at least one prolonged interruption we can see the metrics drop in comparison with short interruptions and the selections are now more frequent than the editions events. This agrees with the results with Mylyn on this point.

As for the observation that activity was overall higher in the middle of the session we do see this for sessions without interruptions. For sessions with interruptions, selection activity appears to be constant over time, while edit activity appears to rise over time with a slight drop at the end.

On a local analysis around the interruptions, we can see a similar interesting behavior of the edition and selections metrics as we observed with Mylyn. In the Figure \ref{fig:all_ints_udc} we show the frequency of editions and selections around the interruptions, specifically 30 minutes before and after. We can see that close to the interruption the frequency of editions and selections drastically drop, and in a radius of around 10 minutes both metrics are below the overall median. This phenomenon is also visible on the same version of this plot with Mylyn, agreeing that the activity of programming gradually decreases before the interruption and takes several minutes to restore the intensity of work above the median. We also see that the edit activity in UDC data is \emph{more} affected by interruptions compared to selection activity: the edits pass below the selections events near the interruption. In contrast to the overall trend, the overall effect appears to be more drastic for UDC data this time.

@@

After this, we separated the interruptions in three categories: those with positive effects afterwards, those a with negative effect and those with no clear effect. Both positive and negative interruptions have a proportion of $44\%$ each of all the interruptions, and the neutrals stand with the $12\%$. All proportions are similar to Mylyn. The results are on the Table \ref{tbl:local_effect_udc}.

@@

The table \ref{tbl:local_effect_udc} show a similar pattern on negative, positive and neutral interruptions in comparison with the results on Mylyn. After a positive interruption both metrics tend to increase at a high rhythm. In contrast, when facing the effects of a negative interruption the metrics take longer to reach the median. We can see that once separated in categories, the effect appear to be even more drastic on UDC data than they appear on Mylyn data.

\section{RQ4: Recovery Time after an Interruption}
\subsection{Edition and selections during recovery time}
The amount of information we have with Usage Data Collector allows us to analyze more specific sections on a programming session. We further investigated the immediate activities after an interruption that we name the recovery time, which is a time period taken by the programmer to resume the interrupted work. We could not do this with Mylyn data, since the process involves a lot of filtering of interruptions, which left us with too few datapoints to analyze.

@@

From those valid interruptions, we split the time after the interruption event into two phases: the first 15 minutes labeled as \textit{Recovery Time} and the rest as \textit{After Recovery Time}. We calculated the edit selection and edit ratio metrics of all the phases, after dividing the Recovery Time into five minute chunks to better visualize the evolution of the metrics.

Considering the results obtained on the local analysis of an interruption, we can conclude that before an interruption it is common to have a reduction of both editions and selections per minute. Therefore, the \textit{After Recovery Time} time was split into two phases as well: one that takes the time where the metrics drop which happens 10 minutes before the next interruption (labeled as  \textit{Before Interruption}), and the time before that, which is the peak of the session after the Recovery Time. The duration of the phase \textit{Before Interruption} was chosen with the conclusions after the local analysis around the interruptions.  We can see on the Figure \ref{fig:box_recovery_udc} the evolution of the metrics through the phases.

@@

The three metrics show a tendency to increase and reach a maximum between the last phase of the \textit{Recovery Time} and the peak of the session. Moreover, the fall of the productivity on the last minutes before the next interruptions agrees with the observations on the local analysis of the time around an interruption. From this we can conclude that the first five minutes after an interruption are the period where productivity is more affected, while the next five minutes are less affected. After 10 minutes, performance appears to be close to the peak.

@@

We can see that after a positive interruption the edit ratio during recovery time is much higher than after a negative interruption, which agrees with the results of the local analysis of interruptions and the observed effects on the metrics. %The reason of this change is probably due to the difference on the activities performed after the interruption according to the effect

@@

To better understand the reasons of the difference between positive and negative interruptions, we did an analysis of the common events during Recovery Time, depending on the nature of the interruption. For this we created a new classification of events on a lower level that takes into consideration what kind of activity is performed by the programmer. The new classification contains the categories described on the Table \ref{tbl:new_events_udc}. Similarly to the first classification, we followed an iterative process involving two of the authors to classify the events in these categories.

\begin{table}[ht!]
\renewcommand{\arraystretch}{1.3}
\caption{New category of events. }
\label{tbl:new_events_udc}
\centering
\begin{tabular}{|p{1.7cm}|p{4 cm}|} 
  \hline 
category & description \\  
  \hline 
    \hline 
clean-build &  Events related to the build sysem (clean, build or run a project) \\
control & Version events events \\
debug & Events executed during debugging sessions  \\
edit-text & Text edition events  \\
file & Management of files (saving, closing, opening, etc.)  \\
find-replace & Search and replacement of text  or code entities \\
high-nav & High-level navigation through classes, windows and artifacts \\
perspective & Change of perspectives in Eclipse  \\
refactoring & Refactoring events \\
search & Text or code search  \\
text-nav & Low-level text navigation \\

@@

We calculated a weighted average of every event based on the average frequency of execution during the first phase of Recovery Time after a positive or negative interruption, weighted by the percentage of users that make use of that event (so that events that are used by very few users do not distort the results). To get a baseline, the same frequency calculation was made for the entire dataset, without distinction of the location of events in the session. We then compared the weighted average changes to have evidence of what kind of events are more often executed after a positive or a negative interruption. Table \ref{tbl:stats_events} shows the results.

@@

We can conclude that after a positive interruption there are more assertive actions with a clear intention like refactoring a class or saving all changes; activities related to program edition. After a negative interruption there are more uncertainty and doubt judging by the activities performed like debugging, searching for text or navigating around classes; activities related to program comprehension. These categories of events agree with our previous hypothesis that positive interruptions are related to information seeking, while negative interruptions are related to actual task switches where context needs to be rebuilt afterwards.

It is noticeable the relation between type of events. For example is natural to think that if there is a lot of debugging events there is also a lot of clean and build events as they are necessary to begin a debugging session. There is also a link between high level navigation and debugging because navigating into a method may take you to another class.

\section{Discussion}
Although they represent the same kind of information both Mylyn and UDC datasets have some peculiarities that make them unique and require different treatments.
